A note on the instability and degeneracy of deep learning models

### 2017-11-06 writing/paper-note/chapters/instab.Rmd
ready?

The above model condition incorporates many standard parameterizations and follows, for instance, whenever $P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y)  = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x, \boldsymbol y \in\mathcal{X}^N$ in a FOES model. For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(rbm)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)). When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(thm:degenFOES}).

@@

```{remark, echo=TRUE}
Under the model condition PSR, Corollary \ref{cor:sign} can also be extended to
cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ (say) are not all changed in sign (e.g., $-\boldsymbol \theta_N$) but, more generally, are instead  altered to another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$ involving a switch in sign only among some dominating model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen. If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then the results of Corollary  \ref{cor:sign} can still hold  with $\boldsymbol \theta_{N}^A$ replacing $-\boldsymbol \theta_N$; as an example of one sufficient condition, if $\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0$ holds in addition to Corollary~1 assumptions, where
$$
G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
$$
represents a standardized form of $\boldsymbol \theta$-model probabilities, then the results of Corollary \ref{cor:sign} apply to $\boldsymbol \theta_{N}^A$ in addition to $-\boldsymbol \theta_N$. As a consequence, an unstable model under $\boldsymbol \theta_N$ can then imply that many more unstable models exist  over a broader spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$, which involves some amount of sign change among components of $\boldsymbol \theta_N$. 
```

@@

    based on $f_{\boldsymbol \theta_N}$ from \@ref(eq:f) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.

@@

In Proposition \@ref(prp:prop2)(iii), the relationships between RBM models with regard to instability, and the effects of different parameter types, follow from the bounds on model instability measures in Proposition \@ref(prp:prop2)(i)-(ii). Generally speaking, all instability in the marginal RBM model for the data $\boldsymbol X$ can be attributed to an excessively large model quantity $\elt$, which predominantly follows when main $\boldsymbol \theta_N^{\mathcal{V}}$ and interaction $\boldsymbol \theta_N^{\mathcal{VH}}$ parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as $|\boldsymbol \theta_N^{\mathcal{V}}|_1$, $\Gam$ or $\Gamc$). For example, for any bounded sequence $|\boldsymbol \theta_N^{\mathcal{H}}|/N$ of hidden parameters, if main visible parameters $\boldsymbol \theta_N^{\mathcal{V}}$ are too extreme ($|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty$), this aspect  will guarantee instability in the visible model ($\elt/N\to \infty$). In fact, the instability measure $\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ for marginal/visible model represents a clearly smaller portion of the  instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ in the joint RBM model, implying that an unstable marginal model (i.e., due to  $\boldsymbol \theta_N^{\mathcal{V}}$, $\boldsymbol \theta_N^{\mathcal{VH}}$) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size $|\boldsymbol \theta_N^{\mathcal{H}}|_1$. For example, while the joint RBM model for $(\boldsymbol X,\boldsymbol H)$ must always be unstable due to a diverging combination of visible and/or interaction parameters ($|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty$ or $\Gam/N\to \infty$) (Proposition \@ref(prp:prop2)(iii.2)), instability for the joint model can also result when the main hidden parameters $\boldsymbol \theta_N^{\mathcal{H}}$ become too large relative to sample size ($|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty$ in Proposition \@ref(prp:prop2)(iii.5)). However, under Proposition \@ref(prp:prop2), the main hidden parameters $\boldsymbol \theta_N^{\mathcal{H}}$  do not necessarily entail a source of instability for the marginal visible model. To explain this distinction, consider a joint model where all parameters related to visibles are zero, $\boldsymbol \theta_N^{\mathcal{V}}= \boldsymbol \theta_N^{\mathcal{VH}}=\boldsymbol 0$, but the hidden-related parameters diverge in sum $|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty$. Here, the explosive behavior among parameters $\boldsymbol \theta_N^{\mathcal{H}}$ induces instability in the joint model for $(\boldsymbol X, \boldsymbol H)$ but the marginal model for $\boldsymbol X$, however, has a perfectly stable (and in fact uniform) distribution in this case. When the hidden parameters are bounded relative to the sample size ($\sup_{N\geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1/N<\infty$), then all instability in both the joint and marginal RBM models can be directly linked to excessively large visible $\boldsymbol \theta_N^{\mathcal{V}}$ and/or interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$ so that features of stability/instability must be the same across both models (Proposition \@ref(prp:prop2)(iii.6)). Hence, to prevent instability in the joint model, the combined magnitudes of all parameters $\theta_N$ must be controlled (cf. Proposition \@ref(prp:prop2)(iii.7)), while a stable visible data model technically results in constraining only the sizes of visible-related parameters $\boldsymbol \theta_N^{\mathcal{V}}$, $\boldsymbol \theta_N^{\mathcal{VH}}$. Nevertheless, because the joint model is often employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well).

@@

Volatility in the probability structure of an unstable model can also hamper efforts to maximize likelihood functions in statistical inference. When a FOES model is unstable along a parameter sequence \(\boldsymbol \theta_N\), the same model can further be unstable along parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
origin (model condition PSR and Corollary \@ref(cor:sign)). This can translate into
potential sensitivity of the likelihood function around zero, and lead to numerical complications in maximizing the objective function. We next provide a discussion of this issue in a way that builds upon and extends related findings by @@schweinberger2011instability, who largely focused on the case of one-parameter exponential models.

>>>>

The above model condition incorporates many standard parameterizations and follows, for instance, whenever $P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y)  = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x, \boldsymbol y \in\mathcal{X}^N$ in a FOES model. For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(rbm)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)). When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(thm:degenFOES).

@@

```{remark, echo=TRUE}
Under the model condition PSR, Corollary \ref{cor:sign} can also be extended to
cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ (say) are not all changed in sign (e.g., $-\boldsymbol \theta_N$) but, more generally, are instead  altered to another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$ involving a switch in sign only among some dominating model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen. If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then the results of Corollary  \ref{cor:sign} can still hold  with $\boldsymbol \theta_{N}^A$ replacing $-\boldsymbol \theta_N$; as an example of one sufficient condition, if $\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0$ holds in addition to Corollary \@ref(cor:sign) assumptions, where
$$
G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
$$
represents a standardized form of $\boldsymbol \theta$-model probabilities, then the results of Corollary \ref{cor:sign} apply to $\boldsymbol \theta_{N}^A$ in addition to $-\boldsymbol \theta_N$. As a consequence, an unstable model under $\boldsymbol \theta_N$ can then imply that many more unstable models exist  over a broader spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$, which involves some amount of sign change among components of $\boldsymbol \theta_N$. 
```

@@

    based on $f_{\boldsymbol \theta_N}$ from (\ref{eq:f}) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.

@@

In Proposition \@ref(prp:prop2)(iii), the relationships between RBM models with regard to instability, and the effects of different parameter types, follow from the bounds on model instability measures in Proposition \@ref(prp:prop2)(i)-(ii). Generally speaking, all instability in the marginal RBM model for the data $\boldsymbol X$ can be attributed to an excessively large model quantity $\elt$, which predominantly follows when main $\boldsymbol \theta_N^{\mathcal{V}}$ and interaction $\boldsymbol \theta_N^{\mathcal{VH}}$ parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as $|\boldsymbol \theta_N^{\mathcal{V}}|_1$, $\Gam$ or $\Gamc$). For example, for any bounded sequence $|\boldsymbol \theta_N^{\mathcal{H}}|/N$ of hidden parameters, if main visible parameters $\boldsymbol \theta_N^{\mathcal{V}}$ are too extreme ($|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty$), this aspect  will guarantee instability in the visible model ($\elt/N\to \infty$). In fact, the instability measure $\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ for marginal/visible model represents a clearly smaller portion of the  instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ in the joint RBM model, implying that an unstable marginal model (i.e., due to  $\boldsymbol \theta_N^{\mathcal{V}}$, $\boldsymbol \theta_N^{\mathcal{VH}}$) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size $|\boldsymbol \theta_N^{\mathcal{H}}|_1$. For example, while the joint RBM model for $(\boldsymbol X,\boldsymbol H)$ must always be unstable due to a diverging combination of visible and/or interaction parameters ($|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty$ or $\Gam/N\to \infty$) (Proposition \@ref(prp:prop2)(iii.2)), instability for the joint model can also result when the main hidden parameters $\boldsymbol \theta_N^{\mathcal{H}}$ become too large relative to sample size ($|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty$ in Proposition \@ref(prp:prop2)(iii.5)). However, under Proposition \@ref(prp:prop2), the main hidden parameters $\boldsymbol \theta_N^{\mathcal{H}}$  do not necessarily entail a source of instability for the marginal visible model. To explain this distinction, consider a joint model where all parameters related to visibles are zero, $\boldsymbol \theta_N^{\mathcal{V}}= \boldsymbol \theta_N^{\mathcal{VH}}=\boldsymbol 0$, but the hidden-related parameters diverge in sum $|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty$. Here, the explosive behavior among parameters $\boldsymbol \theta_N^{\mathcal{H}}$ induces instability in the joint model for $(\boldsymbol X, \boldsymbol H)$ but the marginal model for $\boldsymbol X$, however, has a perfectly stable (and in fact uniform) distribution in this case. When the hidden parameters are bounded relative to the sample size ($\sup_{N\geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1/N<\infty$), then all instability in both the joint and marginal RBM models can be directly linked to excessively large visible $\boldsymbol \theta_N^{\mathcal{V}}$ and/or interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$ so that features of stability/instability must be the same across both models (Proposition \@ref(prp:prop2)(iii.6)). Hence, to prevent instability in the joint model, the combined magnitudes of all parameters $\theta_N$ must be controlled (cf. Proposition \@ref(prp:prop2)(iii.7)), while a stable visible data model technically results in constraining only the sizes of visible-related parameters $\boldsymbol \theta_N^{\mathcal{V}}$, $\boldsymbol \theta_N^{\mathcal{VH}}$. Nevertheless, because the joint model is often employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well). Further evidence of this is seen in the following numerical example. \par

```{r rbm-simul, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, cache.lazy = FALSE}
#libraries -----------------------
source("../resources/code/functions.R")

#functions ----------------------
sample_sphere_unif <- function(space, rep, center, r) {
  vec <- matrix(rnorm(space*rep), nrow = rep)
  r*vec/t(matrix(rep(sqrt(rowSums(vec^2)), each = space), nrow = space))
}

#sample thetas --------------
n <- 100
V <- 9
H <- 5

expand.grid(H = H, V = V, r1 = seq(0.001, 3, length.out = 20), r2 = seq(0.001, 3, length.out = 20)) %>%
  mutate(r1 = r1*(H + V), r2 = r2*(H*V)) %>%
  mutate(N = H + V) %>%
  group_by(H, V, r1, r2, N) %>%
  do(samp = cbind(sample_sphere_unif(.$N, n, 0, .$r1), sample_sphere_unif(.$H*.$V, n, 0, .$r2))) -> grid_sample

expand.grid(H = H, V = V, r1 = seq(0.001, 3, length.out = 20), r2 = seq(0.001, 3, length.out = 20)) %>%
  mutate(r1 = r1*(H + V), r2 = r2*(H*V)) %>%
  mutate(N = H + V) %>%
  group_by(H, V, r1, r2, N) %>%
  do(stat = stats(.$H, .$V, "negative")) -> grid_sample_stat

grid_sample <- inner_join(grid_sample, grid_sample_stat)

grid_sample %>%
  group_by(H, V, N, r1, r2) %>%
  do(g_theta = t(rbind(t(.$samp[[1]]), expected_value(theta = t(.$samp[[1]]), stats = data.matrix(.$stat[[1]]))))) -> grid

res <- inner_join(grid, grid_sample)

#reshape data functions ----------------
elpr <- function(theta, stats) {
  require(dplyr)
  exp(tcrossprod(stats, theta)) %>% data.frame() %>%
    cbind(stats) %>%
    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
    summarise_each(funs(sum), contains("X")) %>%
    ungroup() -> marginalized

  marginalized %>%
    summarise_each(funs(max), contains("X")) %>%
    select(contains("X")) %>%
    data.matrix() -> max_marg

  marginalized %>%
    summarise_each(funs(min), contains("X")) %>%
    select(contains("X")) %>%
    data.matrix() -> min_marg

  t(log(max_marg/min_marg))
}

#instability
res %>%
  group_by(H, V, N, r1, r2) %>%
  do(data.frame(elpr = elpr(.$samp[[1]], .$stat[[1]]))) %>%
  ungroup() %>%
  group_by(H, V, N, r1, r2) %>% 
  summarise(mean_elpr = mean(elpr)) %>%
  mutate(scaled_mean_elpr = mean_elpr/V) %>%
  ungroup() -> elpr_summary

res %>%
  group_by(H, V, N, r1, r2) %>%
  do(data.frame(max_prob = apply(.$samp[[1]], 1, function(row) {
    params <- list(main_visible = row[1:V] %>% data.matrix(),
                   main_hidden = row[(V + 1):(V + H)] %>% data.matrix(),
                   interaction = row[(V + H + 1):length(row)] %>% data.matrix() %>% matrix(H))

    distn <- visible_distn(params) %>% mutate(out_vals = pmap(lst_(paste0("v", 1:V)), c)) %>% select(-(starts_with("v")))

    compare <- expand.grid(out1 = distn$image_id, out2 = distn$image_id) %>%
      left_join(distn, by = c("out1" = "image_id")) %>%
      left_join(distn, by = c("out2" = "image_id")) %>%
      mutate(test = map2_dbl(out_vals.x, out_vals.y, function(a, b) sum(a != b))) %>%
      filter(test == 1) %>%
      mutate(delta = log(prob.x/prob.y))

    max(compare$delta)
    })
    )) %>%
  ungroup() %>%
  group_by(H, V, N, r1, r2) %>% 
  summarise(mean_max_prob = mean(max_prob)) %>%
  ungroup() -> mean_max_prob_diff_summary
```

```{r rbm-plots, fig.show='hold', fig.cap="The sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/N_{\\mathcal{V}}$ (left) and $\\DN(\\boldsymbol \\theta)$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$. As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of these metrics, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
elpr_summary %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(paste("Mean ", frac(ELPR(theta), N[italic(V)]))), low = "yellow", high = "red") +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", N[italic(H)]," + ", N[italic(V)], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", N[italic(H)], "*", N[italic(V)], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom")

# fix labels
mean_max_prob_diff_summary %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = mean_max_prob)) +
  geom_contour(aes(x = r1, y = r2, z = mean_max_prob), colour = "black", bins = 8) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(paste("Mean ", Delta[N](theta))), low = "yellow", high = "red") +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", N[italic(H)]," + ", N[italic(V)], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", N[italic(H)], "*", N[italic(V)], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom") 
```

In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters $\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})$ and interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$) to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})$ and $||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})$ for a RBM with $N_\mathcal{V} = `r V`$ visibles and $N_\mathcal{H} = `r H`$ hiddens. These average magnitudes vary on a grid between `r min(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))` and `r max(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))` with `r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))` breaks, yielding `r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))^2` grid points. At each point in the grid, `r n` vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and `r n` vectors ($\boldsymbol \theta_{interaction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables. These vectors are then paired to create `r n` values of $\boldsymbol \theta_N$ with magnitudes at each point in the grid. The values $\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}$ and $\DN(\boldsymbol \theta_N)$ are then calculated for each $\boldsymbol \theta_N$ and then summarized for each point in the grid using the sample mean. The results of this numerical study are shown in Figure \@ref(fig:rbm-plots). From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by $N_{\mathcal{V}}$ and the the biggest log-probability ratio for a one component change in data outcomes are both increasing away from $\boldsymbol \theta_N = \boldsymbol 0$, further supporting \@ref(prp:prop2)(iii.2 and iii.5).

@@

Volatility in the probability structure of an unstable model can also hamper efforts to maximize likelihood functions in statistical inference. When a FOES model is unstable along a parameter sequence \(\boldsymbol \theta_N\), the same model can further be unstable along parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
origin (model condition PSR and Corollary \@ref(cor:sign)). This can translate into
potential sensitivity of the likelihood function around zero, and lead to numerical complications in maximizing the objective function. We next provide a discussion of this issue in a way that builds upon and extends related findings by @schweinberger2011instability, who largely focused on the case of one-parameter exponential models.
